---
title: "NBA Roster Construction"
author: "Jacob Mortensen"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: rmarkdown::html_vignette
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "..")
```
The goal of this project is to identify synergies and redundancies within NBA 
rosters. Our initial idea is to use a modified version of RAPM that includes
latent factors for different player types as a predictor variable, and then
examine interactions between the player type factors to identify potential
redundancies or synergies within a given roster. 

First we use EFA to identify latent factors for all of the players in our
data that we can then use in our adjusted plus minus model. Note that we
filter out all players who played less than 250 minutes in a season.

# Create per 36 minute statistics
```{r}
library(dplyr)
data("player_stats_2017REG")
# filter out players with less than 250 minutes
players_df <- players %>% 
  group_by(player_id, full_name, weight, height, experience) %>%
  summarise(minutes = sum(minutes),
            field_goals_made = sum(field_goals_made),
            field_goals_attempted = sum(field_goals_attempted),
            three_points_made = sum(three_points_made),
            three_points_attempted = sum(three_points_attempted),
            two_points_made = sum(two_points_made),
            two_points_attempted = sum(two_points_attempted),
            blocked_attempts = sum(blocked_attempts),
            free_throws_made = sum(free_throws_made),
            free_throws_attempted = sum(free_throws_attempted),
            offensive_rebounds = sum(offensive_rebounds),
            defensive_rebounds = sum(defensive_rebounds),
            rebounds = sum(rebounds),
            assists = sum(assists),
            turnovers = sum(turnovers),
            steals = sum(steals),
            blocks = sum(blocks),
            personal_fouls = sum(personal_fouls),
            points = sum(points)) %>%
  ungroup() %>%
  mutate(fg_per36 = 36 * field_goals_made / minutes,
         fga_per36 = 36 * field_goals_attempted / minutes,
         fg_pct = field_goals_made / field_goals_attempted,
         three_pts_per36 = 36 * three_points_made / minutes,
         three_pts_att_per36 = 36 * three_points_attempted / minutes,
         three_pt_pct = three_points_made / three_points_attempted,
         two_pts_per36 = 36 * two_points_made / minutes,
         two_pts_att_per36 = 36 * two_points_attempted / minutes,
         two_pt_pct = two_points_made / two_points_attempted,
         blocked_attempts_per36 = 36 * blocked_attempts / minutes,
         ft_per36 = 36 * free_throws_made / minutes,
         fta_per36 = 36 * free_throws_attempted / minutes,
         ft_pct = free_throws_made / free_throws_attempted,
         or_per36 = 36 * offensive_rebounds / minutes,
         dr_per36 = 36 * defensive_rebounds / minutes,
         reb_per36 = 36 * rebounds / minutes,
         ast_per36 = 36 * assists / minutes,
         to_per36 = 36 * turnovers / minutes,
         ato_ratio = assists / turnovers,
         stl_per36 = 36 * steals / minutes,
         blk_per36 = 36 * blocks / minutes,
         pf_per36 = 36 * personal_fouls / minutes,
         pts_per36 = 36 * points / minutes,
         height = as.numeric(height),
         weight = as.numeric(weight),
         experience = as.numeric(experience))
```

# Exploratory factor analysis on player statistics
Here we try EFA on several different groups of variables to see what makes the most sense.
```{r}
# remove variables that aren't useful for EFA
minutes_threshold <- 250
df1 <- players_df %>%
  filter(minutes > minutes_threshold) %>%
  select(matches("(per36|pct)"), ato_ratio) %>% 
  select(-fg_per36, -fga_per36, -three_pts_per36, -three_pts_att_per36, 
         -two_pts_per36, -two_pts_att_per36, -or_per36, -dr_per36, 
         -ato_ratio, -fg_pct, -ft_per36, -fta_per36)
df1$three_pt_pct[is.nan(df1$three_pt_pct)] <- 0
df1$ft_pct[is.nan(df1$ft_pct)] <- 0
df1 <- scale(df1)
ev <- eigen(cov(df1))
plot(1:length(ev$values), ev$values)
fa1 <- factanal(df1, factors = 5, scores = "regression")
fa1
estimated_vals <- fa1$scores %*% t(fa1$loadings)
clust1 <- hclust(dist(fa1$scores))
grps <- cutree(clust1, k = 7)
players_df %>%
  filter(minutes > minutes_threshold) %>%
  filter(grps == 6) %>%
  select(full_name)
# rind <- sample(1:nrow(df1), size = 1)
# rbind(estimated_vals[rind, ], df1[rind, ])

df2 <- players_df %>%
  filter(minutes > minutes_threshold) %>%
  select(height, weight, matches("per36"), ato_ratio) %>%
  select(-fg_per36, -fga_per36, -ast_per36, -to_per36, -reb_per36, -pts_per36)
df2 <- scale(df2)
ev <- eigen(cov(df2), only.values = T)
plot(1:length(ev$values), ev$values)
fa2 <- factanal(df2, factors = 6, scores = "regression", lower = 0.05)
fa2
evals2 <- fa2$scores %*% t(fa2$loadings)
# rind <- sample(1:nrow(df2), size = 1)
# rbind(evals2[rind, ], df2[rind, ])

df3 <- players_df %>%
  filter(minutes > minutes_threshold) %>%
  select(height, matches("(per36|pct)")) %>%
  select(-fg_per36, -fga_per36, 
         -three_pts_per36, -three_pts_att_per36, -three_pt_pct,
         -two_pts_per36, -two_pts_att_per36, -two_pt_pct,
         -ft_per36, -fta_per36, -or_per36, -dr_per36)
df3$ft_pct[is.nan(df3$ft_pct)] <- 0
df3 <- scale(df3)
ev <- eigen(cov(df3), only.values = T)
plot(1:length(ev$values), ev$values)
fa3 <- factanal(df3, factors = 6, scores = "regression", lower = 0.05)
fa3
evals3 <- fa3$scores %*% t(fa3$loadings)
# rind <- sample(1:nrow(df3), size = 1)
# rbind(evals3[rind, ], df3[rind, ])
```

# Raw adjusted plus minus
```{r}
library(roster)
library(dplyr)
data("all_pbp")
plus_minus_df <- plus_minus(pbp)
# n_before <- nrow(plus_minus_df)
# pls_min_df <- plus_minus_df %>%
#   filter(home_points_per_poss <= 3, away_points_per_poss <= 3)
# n_after <- nrow(pls_min_df)
# cat(n_before - n_after, " lineups removed due to problems with their per_poss stats")
minutes_threshold <- 0
fit <- apm(plus_minus_df, players_df, minutes_threshold = minutes_threshold, weights = T)
betahat <- sort(fit$beta[, 4], decreasing = T)
players_df[match(names(betahat), players_df$player_id), ] %>% select (full_name)
minutes_threshold <- 1000
fit <- apm(plus_minus_df, players_df, minutes_threshold = minutes_threshold, weights = T)
betahat <- sort(fit$beta[, 4], decreasing = T)
players_df[match(names(betahat), players_df$player_id), ] %>% select (full_name)
```


